---
title: "Assignment 1"
author: "David Samson"
date: "August 12, 2015"
output: html_document
---

This project analyzes data collected from a single individual at 5 minute intervals through out the day, during the months of October and November, 2012.  

The basic steps which will be followed will be: 
1. Load the raw data.
2. Create a modified data set to handle missing values.  
3. The following data analysis steps on the raw data:
  + Histogram of the total number of steps per day
  + Median and mean per day
  


## Loading and preprocessing the data
```{r}
library(dplyr)
raw<-read.csv("activity.csv")
summary(raw) # For simple overview of loaded data.  
```
 
## What is mean total number of steps taken per day?
```{r}
grouped<-group_by(raw,date)
dailyTotals <- summarise(grouped,sumSteps=sum(steps),meanSteps=mean(steps,na.rm=T),medianSteps=median(steps,na.rm=T))
```

## Overview plot of the data
breaks <-  as.Date.POSIXct(c("1-Oct-2012","15-Oct-2012","30-Oct-2012","14-Nov-2012","29-Nov-2012"))
p<-ggplot(dailyTotals,aes(x=date, y=sumSteps)) + geom_histogram(stat="identity")+scale_x_continuous(breaks=breaks)

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
